---
title: "interaction"
author: "Xue Zhang"
date: "2023-12-10"
output: html_document
---

```{r}
library(tidyverse)
library(survival)
library(glmnet)
library(pROC)
```


```{r}
clean_data = read_csv("./data/clean_data.csv")
new_data = read_csv("./data/new_data.csv")


```



```{r}
# Fit a Cox proportional hazards model
cox_model = coxph(Surv(time = survival_months, event = status) ~ age + race + marital_status + t_stage +
                    n_stage + x6th_stage + differentiate + grade + a_stage + tumor_size +
                    estrogen_status + progesterone_status + regional_node_examined +
                    regional_node_positive, data = clean_data)

# Summarize the model to see which factors are significant
summary(cox_model)

# You can also look for interactions by adding interaction terms to the model, 
# but this should be done cautiously to avoid overfitting, especially with many covariates
# Example interaction between age and t_stage
cox_model_interaction = coxph(Surv(time = survival_months, event = status) ~ age * t_stage, data = clean_data)

# Summarize the model with interaction
summary(cox_model_interaction)
```





# Potential Interaction terms according to Common Sense

Tumor Size and Stage: The impact of tumor size on survival may vary depending on the stage of the cancer. For example, a large tumor might be more significant in early stages but less so in advanced stages where other factors become more dominant.

Marital Status and Age: The support network impact on survival might be more pronounced for certain age groups, indicating a possible interaction between marital status and age.





# Predicting Risk of Death Using Covariates
Build a predictive model using variables 1-14 as covariates.
```{r}
covariates = clean_data[, 1:14]
outcome = clean_data$status

model_data = data.frame(covariates, outcome)

model = glm(outcome ~ ., data = model_data, family = "binomial")
```

# Identifying Significant Factors and Interactions
To identify significant factors and potential interactions:
```{r}
summary(model)

# race + age
model_interaction = glm(outcome ~ . + race:age, data = model_data, family = "binomial")
summary(model_interaction)
```

**age:race Interaction Term: The coefficient for this term shows how the effect of age on the outcome varies across different race groups. However, this interaction term is not statistically significant (p = 0.210452), suggesting that the effect of age on the outcome does not significantly differ across races in your dataset.**

**Model Comparison: Comparing the AIC of both models (3027.4 vs. 3027.9), the initial model without the interaction term seems to have a slightly better fit, although the difference is very minimal.**


# Evaluating Model Performance and Fairness
To evaluate model performance:
```{r}
# Predicting and evaluating performance
predictions = predict(model, type = "response")

roc_curve = roc(outcome, predictions)

auc(roc_curve)
```

**The AUC value is 0.7455. This metric ranges from 0 to 1, where an AUC of 0.5 suggests no discriminative ability (equivalent to random guessing), and an AUC of 1 indicates perfect discrimination.**

**An AUC of 0.7455 suggests that your model has a good ability to distinguish between the case and control groups. This means that, in general, the model assigns higher probabilities to actual cases than to controls.**

To check the performance across different racial groups and ensure fairness
```{r}
# Splitting data by racial groups
data_white  = clean_data[clean_data$race == "0", ]
data_black_other =  clean_data[clean_data$race %in% c("1", "2"), ]

# Evaluate model separately for these groups
predictions_white = predict(model, newdata = data_white, type = "response")
roc_white = roc(data_white$status, predictions_white)

predictions_black_other = predict(model, newdata = data_black_other, type = "response")
roc_black_other = roc(data_black_other$status, predictions_black_other)

# Compare AUC scores
auc_white = auc(roc_white)
auc_black_other = auc(roc_black_other)


```

# race : age

```{r}
# Convert numerical values to factors with appropriate labels
clean_data$race = factor(clean_data$race,
                         levels = c(0, 1, 2),
                         labels = c("White", "Black", "Other"))

clean_data$age = cut(clean_data$age, 
                        breaks = c(-Inf, 30, 50, Inf),
                        labels = c("Under 30", "30-50", "Over 50"),
                        right = FALSE)

clean_data$status = factor(clean_data$status, 
                            levels = c(0, 1),
                            labels = c("Dead", "Alive"))

# Group and summarize the data
summary_data_a = clean_data |>
  group_by(age, race, status) |>
  summarise(count = n(), .groups = "drop")

# Create a bar plot
ggplot(summary_data_a, aes(x = age, y = count, fill = status)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~race, scales = "free_y") +
  labs(title = "Age and Death Status by Race", 
       x = "Age", 
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        strip.background = element_rect(fill = "lightblue"),
        strip.text.x = element_text(size = 10, face = "bold"))


```


# race: marital status


# Predicting Risk of Death Using Covariates
Build a predictive model using variables 1-14 as covariates.
```{r}
covariates = clean_data[, 1:14]
outcome = clean_data$status

model_data = data.frame(covariates, outcome)

model = glm(outcome ~ ., data = model_data, family = "binomial")
```

# Identifying Significant Factors and Interactions
To identify significant factors and potential interactions:
```{r}
summary(model)

# race + age
race_marital_status_model_interaction = glm(outcome ~ . + race:marital_status, data = model_data, family = "binomial")
summary(race_marital_status_model_interaction)
```

**The interaction raceOther:marital_status is significant, suggesting that the combined effect of being of 'Other' race and marital status is different than their individual effects.**

**Model Comparison: The AIC for this model is 3026, slightly lower than the initial model, suggesting a marginally better fit with the inclusion of the interaction terms.**

```{r}
# Convert numerical values to factors with appropriate labels


clean_data$marital_status = factor(clean_data$marital_status, 
                                    levels = c(0, 1, 2, 3, 4),
                                    labels = c("Married", "Single", "Divorced", "Widowed", "Seperated"))


# Group and summarize the data
summary_data_b = clean_data |>
  group_by(marital_status, race, status) |>
  summarise(count = n(), .groups = "drop")

# Create a bar plot
ggplot(summary_data_b, aes(x = marital_status, y = count, fill = status)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~race, scales = "free_y") +
  labs(title = "Marital Status vs Death Status by Race", 
       x = "Marital Status", 
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        strip.background = element_rect(fill = "lightblue"),
        strip.text.x = element_text(size = 10, face = "bold"))


```

The bar chart illustrates the count of individuals by marital status and their survival status, segmented by race. It's evident that for the White and Other race categories, the majority of individuals are married and alive, while the Black race category has a higher count of single individuals who are alive. Across all race categories, the number of deceased individuals is significantly lower than those alive, with the widowed status showing a higher count of deceased individuals particularly in the White race category.